The benefits of reporting critical effect size values

### 2024-11-26 paper/paperDL.tex
^^

% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}

@@

\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}

@@

   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}

@@

  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}

@@

\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{critical effect size values, statistical significance, fucntions implementation\newline\indent Word count: 3912}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{csquotes}
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother
\usepackage{xcolor}
\newcommand{\brev}{\color{red}}
\newcommand{\erev}{\color{black}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={The benefits of reporting critical effect size values},
  pdfauthor={Ambra Perugini1, Filippo Gambarota1, Enrico Toffalini2, Daniël Lakens3, Massimiliano Pastore1, Livio Finos4, Psicostat1, \& Gianmarco Altoè1},
  pdflang={en-EN},
  pdfkeywords={critical effect size values, statistical significance, fucntions implementation},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\abstract{%
Critical effect size values represent the smallest detectable effect that would reach statistical significance given a certain sample size. These values are highly applicable in various scenarios, as they do not require specifying a plausible effect size of interest. Reporting critical effect size values may be especially useful when adopting NHST (Null Hypothesis Significance Testing) and the sample size has not been planned a priori, or when there is uncertainty about the expected effect size. To assist researchers in calculating critical effect size values, we have developed an R package that allows to report the critical effect size values of model parameters. The package's functions are implemented for group comparisons, correlations, linear regression, and meta-analysis. What we propose could benefit researchers during the planning phase of the study by helping them to understand the limitations and strengths of their research design, and also as a retrospective tool to potentially reframe original interpretations.
}

@@

In the present paper, we suggest that researchers should systematically report critical effect size values when they make inference using statistical significance. The ``critical effect size value(s)'' refers to the smallest statistically significant effect sizes, which depend on the test performed and the characteristics of the sample (Lakens, 2022). To cut to the chase, let us consider a study that tests (using for example a two-tailed test) a bivariate correlation with \(n = 20\) participants and uses \(p < 0.05\) as the threshold for statistical significance: in this case, the smallest significant effects (critical effect size values) are Pearson's \(r = -0.44\) and \(r = 0.44\). As we will argue, critical effect size values can serve as a useful complement to aid in the interpretation of study results, to communicate their relevance, and to help assess their replicability, particularly when statistical significance is relied upon for inference. This is especially important in cases where the sample size was not (or could not) be predetermined based on a theory-driven target effect size, or when statistical power may be difficult to determine or deviates from an optimal level.

Despite longstanding criticisms, Null Hypothesis Significance Testing (NHST) remains a prominent approach for statistical inference in science (Cohen, 1994; Gigerenzer et al., 2004; Krueger, 2001). Over the past decades, numerous proposals have emerged to enhance inference based on this approach. These include conducting power analyses to determine sample size \textit{a priori} based on theory-informed target effects, and reporting effect sizes and their confidence intervals, accompanied by pertinent related comments, rather than drawing dichotomous conclusions solely based on statistical significance (Transue, 2019). Even for the most skeptical, there are suggestions that NHST may possess merits of its own. For instance, Wilson et al. (2020) propose that it could serve as a useful preliminary filter utilized by ``original science'' for screening potentially interesting effects, which should subsequently be validated by ``replication science''.

In psychology, statistical power is often severely limited for medium to small effect sizes (Szucs \& Ioannidis, 2017), which are generally expected to represent true replicable findings. In fact, resource constraints have been authoritatively listed among acceptable justifications for sample size (Lakens, 2022), and this scenario is arguably widespread in psychology. However, limited power makes it challenging to distinguish signal from noise, and on average it results in an overestimation of effect sizes when relying on statistical significance for inference (Altoè et al., 2020). On the other hand, very large samples are occasionally available in psychological science, albeit less frequently. However, as some have argued, excessively large sample sizes risk creating an ``everything-is-significant'' scenario, where researchers report and discuss practically negligible effects solely because they are statistically significant, regardless of theoretical relevance or whether they merely reflect minor procedural artifacts (Wilson et al., 2020). Thus, while statistical significance can serve as a potentially useful filter, relying solely on it for inference may introduce risks in interpretation. In such cases, we contend that reporting critical effect size values can help put into perspective and convey the relevance of the results.

There are two clear benefits to reporting the critical effect size value(s) for a corresponding test. First, when sample sizes are small, the critical effect size values inform readers about whether the effect sizes that could lead to rejecting the null hypothesis are in line with realistic expectations. If the sample size is small and only very large effects would yield a statistically significant result, and the underlying mechanism that is examined is unlikely to lead to such large effect sizes, in these cases, researchers will realize they are not able to collect sufficient data to perform a meaningful test. An a-priori power analysis would typically lead to a similar conclusion, but reporting critical effect size values focuses the attention more strongly on which effect sizes are reasonable to expect. Second, in large studies the critical effect size values will make it clear that trivially small effect sizes will be statistically significant. This will focus the attention of researchers on the difference between statistical significance and practical significance. This is especially important in correlational studies on large samples in psychology, where small uncontrolled sources of statistical noise may simultaneously affect and create some shared variance across otherwise unrelated variables, a phenomenon controversially labelled as the `crud factor' (Orben \& Lakens, 2020).

Let us consider the following two scenarios as examples. First, imagine researchers conducting a study involving a between-group comparison. Due to severe resource constraints, they are only able to collect a limited sample size of \(n = 30\) (\(n = 15\) per group). No statistically significant effect is detected with \(p < 0.05\). Given their prior expectation that the effect of interest may not be large, they acknowledge that their study was likely underpowered, though uncertain to some extent. Subsequently, they compute the critical effect size values, revealing Cohen's \(d = -0.75\) and \(d = 0.75\). This indicates that any estimated \textbar Cohen's \(d| < 0.75\) will certainly fail to reach statistical significance. By reporting these critical effect size values, the researchers transparently convey that estimated effects up to a medium-to-large magnitude will consistently fall short of significance. In essence, this scenario exemplifies a version of the ``winner's curse'' (Hedges, 1984), made explicit. In this context, the winner's curse indicates the tendency for a statistically significant initial finding (i.e., a ``winner'') to be associated with an overestimated effect size which will likely be deflated in subsequent replications (i.e., the ``curse''), especially if such initial finding was obtained with a small sample. Knowing from the critical effect size values that a specific result must necessarily be associated with either an estimated large effect size or be non-significant, readers are, in principle, prevented from both over-interpreting the magnitude of an effect size and from equating lack of statistical significance with lack of an effect of interest.

In a second scenario, imagine researchers who gain access to a very large archival dataset (\(n = 5,000\)) and decide to explore some bivariate correlations. With such a large sample size, researchers are not constrained by sample size limitations and could instead focus on effect sizes. However, they legitimately opt to use statistical significance of a two-tailed test as their primary criterion for whether to comment on effects in their further discussions. In this context, the critical effect size values for significance are determined to be correlations of \(r = \pm 0.03\). This implies that practically any effect differing from zero, regardless of its small magnitude, may easily attain statistical significance. It is worth noting that such small effects may potentially reflect just minor artifacts (Wilson et al., 2020), such as subtle experimenter effects or slight non-independence among observations, even in meticulously designed studies. While readers may be tempted to interpret any reported effects simply because they attain statistical significance, signaling the critical effect size value level beforehand can serve as a clear warning that the significance filter might be overly permissive, thus urging interpretive caution.

A similar scenario may arise in meta-analysis. Despite potential loss of precision due to substantial heterogeneity across effect sizes in different studies, meta-analyses typically synthesize a large amount of evidence, leading even very small average effect sizes to reach significance. While the focus of meta-analysis is generally on estimating effect sizes with uncertainty, statistical significance is routinely reported and interpreted. Thus, in meta-analysis, reporting the critical effect size values can caution readers that the mere attainment of statistical significance by the average effect does not automatically ensure scientific or real-life relevance.

To provide more tangible illustrations of the practical applications of critical effect size values, let us examine two real-world instances drawn from published research. The first one features a small sample and has been targeted in the series of replications by the Open Science Collaboration (Collaboration, 2015), while the second one presents with a very large sample and has already attracted attention in the scientific community. The first case pertains to ``Study 5'' conducted by McCrea (2008). This experiment compared two groups based on their performance percentage in a math test following a preliminary practice session with feedback of failure. One group was exposed to failure-excusing, self-handicapping thoughts, while both groups underwent the math test afterward. With a modest sample size of 28 participants split into two groups (13 and 15 participants, respectively), a directional one-tailed t-test was conducted on the percentage of correct answers to compare the groups. The resulting effect was statistically significant, \(t(26) = 1.87\), \(p\) (one-tailed) \(< .05\). Based on the reported data, Cohen's \(d = 0.736\), indicating a considerable effect size. However, the critical value for Cohen's \(d\) in this context was \(0.645\), suggesting that, given this sample size, statistical significance is achieved only for estimated effects of substantial magnitude. Subsequent replication efforts (Collaboration, 2015) reaffirmed the significance of the effect, and used a somewhat larger sample (\(n = 61\)), \(t(59)=2.325\). The estimated effect size was somehow lower than that originally reported, however, \(d = 0.605\) (here, critical d = 0.428). The second scenario involves the study by Kramer et al. (2014), which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of \(n = 689'003\), the researchers observed a significant effect worsened emotional states when positive posts were reduced, \(t(310,044) = −5.63\), \(p < 0.001\) (non-directional test). Based on this and other results, the authors conclude that emotions expressed by other users on Facebook influence our own emotions. Despite the statistical significance, however, the critical values for Cohen's \(d\) in this case are \(-0.006\) and \(+0.006\), emphasizing the importance of considering effect size beyond mere significance. This prompts reflection on whether an effect of such magnitude on such a large sample (actual \(d = 0.02\)) is truly meaningful in individuals' lives.

\section{How to Compute Critical Effect Size Values}\label{how-to-compute-critical-effect-size-values}

In this section, we provide guidance and formulas for computing critical effect size values in general, with examples for frequently encountered effect sizes including Standardized Mean Differences (Cohen's \(d\)), correlations (Pearson's r), and raw and standardized coefficients in linear models. These formulas have been incorporated into R functions of the package \texttt{criticalESvalue} accessible at: \url{https://github.com/psicostat/criticalESvalue}, and are elaborated upon in the subsequent section.

\subsection{t-test}\label{t-test}

For the t-test we considered the two-sample, paired and one-sample tests. As a general approach the \(t\) statistics is computed as reported in Equation \eqref{eq:eq-gen-t}.

@@

Where \(b\) is the unstandardized effect size that depends on the type of test. For example, in the two sample t-test or the single mean for a one sample t-test. The denominator is the standard error of the numerator that depends on the sample size and the sample(s) variance.

Similarly, Equation \eqref{eq:eq-gen-d} formalize a general form of the effect size.

@@

Where \(b\) is still the unstandardized effect size and \(s\) is the standardization term. For example, in the two-sample case, \(s\) is the pooled standard deviation between the two samples or the standard deviation of the differences for paired samples.

\color{red}

@@

NOT SURE TO KEEP THIS!

@@

\color{black}

\color{red}

\subsubsection{One sample t-test}\label{one-sample-t-test}

The easiest case is the one-sample t-test. \(b\) is the sample mean and \(s\) is the sample standard deviation and \(n\) is the sample size. \(SE_b = \frac{s}{\sqrt{n}}\) and the \(t_c\) is based on \(n - 1\) degrees of freedom. Thus \(b_c = t_c \times SE_b\) and \(d_c = \frac{b_c}{s}\). Equivalently, \(d_c = t_c \times \frac{1}{\sqrt{n}}\)

\color{black}

\subsubsection{Two-sample t-test}\label{two-sample-t-test}

@@

\subsubsection{Paired sample t-test}\label{paired-sample-t-test}

@@

\subsubsection{Hedges's correction}\label{hedgess-correction}

@@

\subsection{Correlation Test}\label{correlation-test}

@@

\subsection{Linear Regression}\label{linear-regression}

Hypothesis testing on regression coefficients (e.g., using the \texttt{lm} function in R) is performed using the so-called Wald test. The test can be considered a one-sample t-test where the \(t\) value is calculated dividing the regression parameter \(\beta_0, \dots, \beta_j\) by its standard error \(SE_{\beta_j}\). The critical \(t\) value is calculated from a \(t\) distribution using \(n - p - 1\) degrees of freedom where \(n\) is the number of observations and \(p\) is the number of coefficients beyond the intercept. We can substitute \(b\) with \(\beta_j\) and \(SE_{b}\) with \(SE_{\beta_j}\) in Equation \eqref{eq:eq-gen-crit} to find the critical unstandardized regression coefficient (\(\beta_{c_j} = t_c \times SE_{\beta_j}\)). There is no an unique way to calculate the standardized version of the critical regression coefficient. The main reason is that in regression modelling we can include categorical and numerical independent variables with main effects and interactions. For example, Groß and Möller (2023) proposed a way to calculate a generalized Cohen's \(d\) from linear regression with or without the presence of other variables dividing the estimated parameter with the residual standard deviation. Gelman (2008) proposed to standardize using two standard deviations because when both numeric and categorical variables are included the coefficients are not on the same metric. Another complication is that in some cases, researchers standardize the dependent variable, the independent variables or both. In general, also the unstandardized critical regression coefficient can be usually easily intepreted. For example, assuming we are regressing some scores on a cognitive tests using the age of participant. The critical regression coefficient is the minimum increase in the cognitive test for a unit age increase that would be significant.

\subsection{Meta-analysis}\label{meta-analysis}

@@

\subsection{Other models}\label{other-models}

Despite we discussed only linear models, the same approach could be applied to other types of models such as generalized linear models. In fact, we simply need to multiply the critical value of the chosen distribution (e.g., \(t\) or \(z\)) by the standard error of the regression coefficient. For example, in a logistic regression with a binary predictor the estimated regression coefficient is the (log) odds ratio. We can obtain the critical odds ratio by multiplying the critical \(z\) statistics (the default in R with the \texttt{glm} function) with the standard error of the regression coefficient.

\section{Examples in R}\label{examples-in-r}

@@

\section{Discussion}\label{discussion}

@@

\section{References}\label{references}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-altoe2020enhancing}
Altoè, G., Bertoldo, G., Zandonella Callegher, C., Toffalini, E., Calcagnı̀, A., Finos, L., \& Pastore, M. (2020). Enhancing statistical inference in psychological research via prospective and retrospective design analysis. \emph{Frontiers in Psychology}, \emph{10}, 499756.

\bibitem[\citeproctext]{ref-Borenstein2009-mo}
Borenstein, M., Hedges, L. V., Higgins, J. P. T., \& Rothstein, H. R. (2009). \emph{Introduction to {Meta-Analysis}}. \url{https://doi.org/10.1002/9780470743386}

\bibitem[\citeproctext]{ref-Bowley1928-yt}
Bowley, A. L. (1928). The standard deviation of the correlation coefficient. \emph{Journal of the American Statistical Association}, \emph{23}, 31. \url{https://doi.org/10.2307/2277400}

\bibitem[\citeproctext]{ref-cohen1994}
Cohen, J. (1994). The earth is round (p \textless{} .05). \emph{American Psychologist}, \emph{49}(12), 997--1003.

\bibitem[\citeproctext]{ref-open2015estimating}
Collaboration, O. S. (2015). Estimating the reproducibility of psychological science. \emph{Science}, \emph{349}(6251), aac4716.

>>>>

@@

\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs

@@

  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}

@@

  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}

@@

\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{critical effect size values, statistical significance, hypothesis testing, effect sizes\newline\indent Word count: 3912}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{csquotes}
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={The benefits of reporting critical effect size values},
  pdfauthor={Ambra Perugini1, Filippo Gambarota1, Enrico Toffalini2, Daniël Lakens3, Massimiliano Pastore1, Livio Finos4, Psicostat1, \& Gianmarco Altoè1},
  pdflang={en-EN},
  pdfkeywords={critical effect size values, statistical significance, hypothesis testing, effect sizes},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\abstract{%
Critical effect size values represent the smallest detectable effect that can reach statistical significance given a specific sample size and alpha level. It can be useful to calculate the critical effect size when designing a study, and evaluate whether such effects are plausible. Reporting critical effect size values may be useful when the sample size has not been planned a priori, or when there is uncertainty about the expected sample size that can be collected, when researchers plan to analyze the data with a statistical hypothesis test. To assist researchers in calculating critical effect size values we have developed an R package that allows researchers to report critical effect size values for group comparisons, correlations, linear regressions, and meta-analyses. Reflecting on critical effect size values could benefit researchers during the planning phase of the study by helping them to understand the limitations of their research design. Critical effect size values are also useful when evaluating studies performed by other researchers when a-priori power analyses were not performed, especially when non-significant results were observed.
}

@@

The ``critical effect size'' refers to the smallest effect size that can be statistically significant, given the test performed, the sample size, and the alpha level (Lakens, 2022). Consider a study that tests a bivariate correlation with \(n = 20\) participants and uses \(\alpha < 0.05\) as the threshold for statistical significance in a two-sided test. The critical effect size values for this test are Pearson's \(r = -0.44\) and \(r = 0.44\), which means that the statistical test will yield a significant result for effects smaller than \(r = -0.44\) or larger than \(r = 0.44\), while observed effect sizes between these two values will lead to a statistically non-significant result. In the present paper, we suggest that researchers should systematically report critical effect size values when they make inference use significance levels in hypothesis tests to make claims about the presence or absence of effects. Critical effect size values can serve as a useful complement interpret study results, communicate the informational value of a study design, and contextualizes the interporetation of non-signiicant results, particularly when statistical significance is relied upon for inference. Reporting critical effect size values is especially important when the sample size was not (or could not) be predetermined on the basis of a smallest effect size of interest.

Null hypothesis significance testing (NHST) remains the most prominent approach for statistical inference in science, even though there are widespread concerns about the misuse of hypothesis tests (Cohen, 1994; Gigerenzer et al., 2004; Krueger, 2001). Over the past decades, numerous proposals have emerged to improve the use of hypothesis testing. These include complementing hypothesis tests with effect sizes and their confidence intervals, preregistering hypotheses before the data is collected, and conducting power analyses to determine the sample size \textit{a priori} based on a smallest effect size of interest. Researchers often publish studies with low power for medium to small effect sizes (Szucs \& Ioannidis, 2017), with the main justification for the sample size being resource limitations (Lakens, 2022). However, low power makes it challenging to distinguish signal from noise, and combined with the selective reporting of statistically significant results in the published literature leads to an overestimation of effect sizes (Altoè et al., 2020). Certain types of research questions can be studied by relying on online data collection, which has made it cheaper and more feasible to collect very large samples. Studies with very large samples are well-powered to detect even small effect sizes, but they also require researchers to carefully consider the possibility that an effect might be statistically significant, but practically insignificant.

We believe there are two clear benefits to reporting critical effect size values for a corresponding test. First, when sample sizes are small, the critical effect size values inform readers about whether the effect sizes that could lead to rejecting the null hypothesis are in line with realistic expectations. If the sample size is small and only very large effects would yield a statistically significant result, and the underlying mechanism that is examined is unlikely to lead to such large effect sizes, researchers will realize they are not able to collect sufficient data to perform an informative hypothesis test. An a-priori power analysis would typically lead to a similar conclusion, but reporting critical effect size values will focus the attention more strongly on which effect sizes are reasonable to expect. Second, when sample sizes are very large, the critical effect size values will make it clear that trivially small effect sizes will be statistically significant. This will focus the attention of researchers on the difference between statistical significance and practical significance, and raises awareness of the importance to interpret the size of the effect. This is especially important in correlational studies with large sample sizes in psychology, where systematic but uncontrolled sources of variability may lead to small but theoretically meaningless non-zero effects, a phenomenon referred to as the `crud factor' (Orben \& Lakens, 2020).

Let us consider the following two scenarios as examples. First, imagine researchers who conduct a study involving a between-group comparison. Due to severe resource constraints, they are only able to collect a limited sample size of \(n = 30\) (\(n = 15\) per group). The authors did not perform an a-priori power analysis, and their sample size justification is purely based on how many participants they can collect. The statistical tests yields a non-significant result, \(p > 0.05\). Given their prior expectation that the effect of interest may not be large, they acknowledge that their study was likely underpowered, though without a theoretically expected effect size, it remains uncertain how low the statistical power is. Subsequently, they compute the critical effect size values, revealing Cohen's \(d = -0.75\) and \(d = 0.75\). This indicates that any observed Cohen's \(d\) between -0.75 and 0.75 will certainly fail to reach statistical significance. By reporting these critical effect size values the researchers transparently convey that estimated effects much larger than most effects in the psychological literature will always fall short of significance. This means that the hypothesis test will not be informative. Non-significant effects are likely, even if there is a true effect of interest. If a significant effect is observed, but most effects in a field are smaller than a study can detect, it is probable the effect size is overestimated (Hedges, 1984), and this should be taken into account when interpreting the effect size.

In a second scenario, imagine researchers who gain access to a very large archival dataset (\(n = 5,000\)) and decide to explore bivariate correlations bwetween variables. With such a large sample size, very small correlations will reach statistical significance, and the critical effect size values for a two-sided test are \(r = \pm 0.03\). It is worth noting that such small effects may potentially reflect just minor artifacts (Wilson et al., 2020), such as subtle experimenter effects or slight non-independence among observations, even in meticulously designed studies. A similar scenario may arise in meta-analysis. Despite potential loss of precision due to substantial heterogeneity across effect sizes in different studies, meta-analyses typically synthesize a large amount of evidence, and as a consequence, even very small average effect sizes can reach significance. While the focus of meta-analysis is generally on estimating effect sizes with uncertainty, statistical significance is routinely reported and interpreted. Signaling the critical effect size values beforehand can serve as a clear warning that statistically significant results should not automatically be interpreted as practically significant, thus urging caution when interpreting the results.

To provide more tangible illustrations of the practical applications of critical effect size values, let us examine two real-world instances drawn from published research. The first one features a small sample and has been targeted in the series of replications by the Open Science Collaboration (Collaboration, 2015), while the second study had a very large sample, and has attracted attention of the scientific community. The first case pertains to ``Study 5'' conducted by McCrea (2008). This experiment compared two groups based on their performance percentage in a math test following a preliminary practice session with feedback of failure. One group was exposed to failure-excusing, self-handicapping thoughts, while both groups underwent the math test afterward. With a modest sample size of 28 participants split into two groups (13 and 15 participants, respectively), a directional one-tailed t-test was conducted on the percentage of correct answers to compare the groups. The resulting effect was statistically significant, \(t(26) = 1.87\), \(p\) (one-tailed) \(< .05\). Based on the reported data, Cohen's \(d = 0.736\), the findings indicated a considerable effect size. However, the critical value for Cohen's \(d\) in this context was \(0.645\), suggesting that, given this sample size, statistical significance is achieved only for estimated effects of substantial magnitude. As the study was not a Registered Report, selective reporting of statistically significant findings was common at the time of publication, and most effects in psychology are substantially smaller, the effect size estimate is likely upwardly biased. Subsequent replication efforts (Collaboration, 2015) reaffirmed the significance of the effect, and used a somewhat larger sample (\(n = 61\)), \(t(59)=2.325\). The estimated effect size was somehow lower than that originally reported, however, \(d = 0.605\) (here, critical d = 0.428). THIS IS NOT A GOOD EXAMPLE, BECAUSE THESE TWO EFFECT SIZE ESTIMATES DO NOT DIFFER STATISTICALLY FROM EACH OTHER, SO WE CAN NOT SAY THE TWO EFFECT SIZES DIFFER! THERE SHOULD BE OTHER EXAMPLES IN THE LITERATURE, MAYBE FROM REGISTERED REPLICATION REPORTS.

The second scenario involves the study by Kramer et al. (2014), which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of \(n = 689.003\), the researchers observed a statistically significant increase in the number of negative words typed by participants when positive posts were reduced in their timeline, \(t(310,044) = −5.63\), \(p < 0.001\) (non-directional test). Based on this and other results from the study, the authors conclude that emotions expressed by other users on Facebook influence our own emotions {[}CHANGE A BIT, STAY CLOSER TO WHAT THE RESEARCHERS SAY, MAYBE QUOTE THEM DIRECTLY{]}. The critical values for Cohen's \(d\) in this experimental design are \(-0.006\) and \(+0.006\). It should have been clear from the outset that the researchers should have considered the effect size beyond merely testing for the statistical significancee. The critical effect size should prompt reflection on which effect sizes are truly meaningful in this context, and whether the observed effect size (\(d = 0.02\)) is theoretically or practically interesting.

\hypertarget{how-to-compute-critical-effect-size-values}{%
\section{How to Compute Critical Effect Size Values}\label{how-to-compute-critical-effect-size-values}}

In this section, we provide guidance and formulas for computing critical effect size values in general {[}BUT WE FOCUS ON STANDARDIZED, NOT RAW, EFFECT SIZES - SO WE EITHER GIVE A REASON FOR THIS, OR WE ALSO ADD RAW EFFECTS{]}, with examples for frequently encountered effect sizes including Standardized Mean Differences (Cohen's \(d\)), correlations (Pearson's \(r\)), and raw and standardized coefficients in linear models. These formulas have been incorporated into R functions of the package \texttt{criticalESvalue} accessible at: \url{https://github.com/psicostat/criticalESvalue}, and are elaborated upon in the subsequent section.

\hypertarget{t-test}{%
\subsection{t-test}\label{t-test}}

For the t-test we considered the two-sample, paired, and one-sample tests. As a general approach the \(t\) statistic is computed as reported in Equation \eqref{eq:eq-gen-t}.

@@

Where \(b\) is the unstandardized effect size that depends on the type of test. {[}THE NEXT SENTENCE NEEDS TO BE REWRITTEN{]} For example, in the two sample \(t\)-test or the single mean for a one sample \(t\)-test. The denominator is the standard error of the numerator that depends on the sample size and the sample(s) variance.

Similarly, Equation \eqref{eq:eq-gen-d} {[}REWRITE THE REST OF THIS SENTENCE{]} formalize a general form of the effect size.

@@

Where \(b\) is the unstandardized effect size and \(s\) is the standardization term. For example, in the two-sample case, \(s\) is the pooled standard deviation between the two samples, while in the paired samples case, \(s\) is the standard deviation of the differences for paired samples.

@@

NOT SURE TO KEEP THIS! {[}DL: I THINK IT IS MAYBE FOR AN APPENDIX{]}

@@

\hypertarget{one-sample-t-test}{%
\subsubsection{One sample t-test}\label{one-sample-t-test}}

For the one-sample \(t\)-test \(b\) is the sample mean and \(s\) is the sample standard deviation and \(n\) is the sample size. \(SE_b = \frac{s}{\sqrt{n}}\) and the \(t_c\) is based on \(n - 1\) degrees of freedom. Thus \(b_c = t_c \times SE_b\) and \(d_c = \frac{b_c}{s}\). Equivalently, \(d_c = t_c \times \frac{1}{\sqrt{n}}\).

\hypertarget{two-sample-t-test}{%
\subsubsection{Two-sample t-test}\label{two-sample-t-test}}

@@

\hypertarget{paired-sample-t-test}{%
\subsubsection{Paired sample t-test}\label{paired-sample-t-test}}

@@

\hypertarget{hedgess-correction}{%
\subsubsection{Hedges's correction}\label{hedgess-correction}}

@@

\hypertarget{correlation-test}{%
\subsection{Correlation Test}\label{correlation-test}}

@@

\hypertarget{linear-regression}{%
\subsection{Linear Regression}\label{linear-regression}}

Hypothesis testing on regression coefficients (e.g., using the \texttt{lm} function in R) is performed using the so-called Wald test. The test can be considered a one-sample t-test where the \(t\) value is calculated dividing the regression parameter \(\beta_0, \dots, \beta_j\) by its standard error \(SE_{\beta_j}\). The critical \(t\) value is calculated from a \(t\) distribution using \(n - p - 1\) degrees of freedom where \(n\) is the number of observations and \(p\) is the number of coefficients beyond the intercept. We can substitute \(b\) with \(\beta_j\) and \(SE_{b}\) with \(SE_{\beta_j}\) in Equation \eqref{eq:eq-gen-crit} to find the critical unstandardized regression coefficient (\(\beta_{c_j} = t_c \times SE_{\beta_j}\)). There is no an unique way to calculate the standardized version of the critical regression coefficient. The main reason is that in regression modelling we can include categorical and numerical independent variables with main effects and interactions. For example, Groß and Möller (2023) proposed a way to calculate a generalized Cohen's \(d\) from linear regression with or without the presence of other variables dividing the estimated parameter with the residual standard deviation. Gelman (2008) proposed to standardize using two standard deviations because when both numeric and categorical variables are included the coefficients are not on the same metric. Another complication is that in some cases, researchers standardize the dependent variable, the independent variables or both. In general, the unstandardized critical regression coefficient can be usually also be easily interpreted. For example, assuming we are regressing some scores on a cognitive tests using the age of participant,the critical regression coefficient is the minimum increase in the cognitive test for a unit age increase that would be significant.

\hypertarget{meta-analysis}{%
\subsection{Meta-analysis}\label{meta-analysis}}

@@

\hypertarget{other-models}{%
\subsection{Other models}\label{other-models}}

Despite the fact that we discussed only linear models, the same approach could be applied to other types of models such as generalized linear models. In fact, we simply need to multiply the critical value of the chosen distribution (e.g., \(t\) or \(z\)) by the standard error of the regression coefficient. For example, in a logistic regression with a binary predictor the estimated regression coefficient is the (log) odds ratio. We can obtain the critical odds ratio by multiplying the critical \(z\) statistics (the default in R with the \texttt{glm} function) with the standard error of the regression coefficient.

\hypertarget{examples-in-r}{%
\section{Examples in R}\label{examples-in-r}}

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

@@

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-altoe2020enhancing}{}}%
Altoè, G., Bertoldo, G., Zandonella Callegher, C., Toffalini, E., Calcagnı̀, A., Finos, L., \& Pastore, M. (2020). Enhancing statistical inference in psychological research via prospective and retrospective design analysis. \emph{Frontiers in Psychology}, \emph{10}, 499756.

\leavevmode\vadjust pre{\hypertarget{ref-Borenstein2009-mo}{}}%
Borenstein, M., Hedges, L. V., Higgins, J. P. T., \& Rothstein, H. R. (2009). \emph{Introduction to {Meta-Analysis}}. \url{https://doi.org/10.1002/9780470743386}

\leavevmode\vadjust pre{\hypertarget{ref-Bowley1928-yt}{}}%
Bowley, A. L. (1928). The standard deviation of the correlation coefficient. \emph{Journal of the American Statistical Association}, \emph{23}, 31. \url{https://doi.org/10.2307/2277400}

\leavevmode\vadjust pre{\hypertarget{ref-cohen1994}{}}%
Cohen, J. (1994). The earth is round (p \textless{} .05). \emph{American Psychologist}, \emph{49}(12), 997--1003.
